Solving an Exam Scheduling Problem Using a Genetic Algorithm

### 2009-12-02 Paper/paper.tex
Paper update

  A schedule, then, is a mapping from courses to rooms at times, 
  which we can express like this:

  \[ C \rightarrow \left\{(r,\ a,\ b)\ |\ r\in R,\ a\in \{1..d\},\ b\in \{1..t\}\right\} \]

  Typically, with relatively loose constraints on the number of rooms and 
  particular schedules of students, it is not difficult to find some consistent
  schedule; that is, one where no student is asked to write 2 exams
  simultaneously and no room has 2 exams occur in it simultaneously.

@@

  The timetable that a student $s \in S$ has under some particular schedule 
  $K$ may be represented in this way:

  \[ TT_s(K) = \left\{(a, b)\ |\ \exists r \in R, \exists c \in L_s, K(c)=(r, a, b) \right\} \]

  which may, in general, be a multiset.

  Similarly, the timetable for a room $r \in R$ is

  \[ TT_r(K)=\left\{(a, b)\ |\ \exists c\in C,\ K(c)=(r, a, b)\right\} \]

  We define two quality functions, mapping schedules to real numbers in [0,1],
  one for students and one for rooms. These are meant to capture how much they
  "like" the schedule under consideration. (For instance, a student will rate
  poorly any schedule where she has 2 consecutive exams, and very poorly any 
  schedule which expects her to take two exams at the same time. A room will 
  rate poorly a schedule that assigns two different exams to be held in it at 
  the same moment).

  \[ Q_s(K)=q_{student}\left(TT_s(K)\right):\textit{ Schedules}\rightarrow [0,\ 1] \]
  \[ Q_r(K)=q_{room}\left(TT_r(K)\right):\textit{ Schedules}\rightarrow [0,\ 1] \]

  We will consider later how to define these functions in a reasonable way.

  The quality of a schedule, then, is given by

  \[ Q(K)= \frac{\left(w_s\left(\sum_{s \in S}{\frac{Q_s(K)}{|S|}}\right) +
     w_r\left(\sum_{r\in R}{\frac{Q_r(K)}{|R|}} \right)\right)}{(w_s+w_r)} \]

  where $w_s$ and $w_r$ are the weightings we can use to indicate that we 
  care somewhat more about students preferences than rooms, giving a quality
  (or fitness, in the terminology of genetic algorithms) in the range [0,1] for
  any schedule.

@@

\section{Approach and Implementation}
  \subsection{Differences From Natural Exam Scheduling Problems}
    The problem we have constructed here is a compromise between more mathematically
    pure problems with the same sort of properties (which might ignore the rooms' 
    constraints completely) and the problem that must be solved in the real world 
    when scheduling thousands of students, which has some important differences from 
    this problem. For example, we don't deal with constraints like room capacity or 
    exams with varying length.

@@

  \subsection{Fitness Function}
    In order to assign each schedule a fitness, we need to define the functions
    $q_{student}$ and $q_{room}$. They need to capture the idea of determining 
    how much each student or room likes their given timetable. We do this by giving 
    penalties to schedules when they have properties that are disliked by the 
    student or room, and then using the formula

    \[ q(TT)=\frac{1}{1+penalty_{TT}} \]

    This function will be in the desired range of [0,1].

@@

    \subsubsection{Representation}
      We use the naive representation of schedules. A schedule contains lists of size $|C|$
      containing the room in which each course's exam is to occur and the timing
      (corresponding to a pair or integers representing the day and time) of each course's
      exam.

      We will see that this representation is far from ideal, but there is no clear 
      alternative.

@@

      It can be seen that this method is completely deterministic given a starting schedule,
      and will always return a local maximum

    \subsubsection{Mutate Search}
      Mutate Search starts by generating a random schedule. It makes an alternative schedule
      (by using the same mutate operator made for the genetic algorithm) with 1\% of the
      data randomized and chooses which of the schedules is better. This operation is
      repeated until it has performed as many evaluations as it is allowed to.

      This method can be changed to true simulated annealing by varying the degree of
      modification made to the current best schedule with the number of evaluations
      remaining. Not being the focus of our investigation, little effort was made to 
      optimize the parameters. 1\% does seem to give better performance than other nearby 
      values, though we did not collect data to substantiate this. Finding ideal parameters 
      for simulated annealing on this problem would make for a project-sized inquiry in itself.

@@

    The following table characterizes the problem instances that we used ($\#$Maj is the 
    number of different majors used and \textsc{SSS} is an approximation of the search space size):    
    \definecolor{light-gray}{gray}{0.87}
    \rowcolors{1}{light-gray}{white}
    \begin{center}
    \begin{small}
    \begin{tabular}{*{7}{c}}
    \hline 
 	  	& S 	& M 	& L 	& $S_{Cor}$ & $M_{Cor}$ & $L_{Cor}$ \\ 
    \hline \hline
     d 		& 7 	& 10	& 7 	& 7 	& 10 	& 7 	\\ 
     t 		& 5		& 8 	& 8 	& 5 	& 8 	& 8 	\\ 
     $|C|$	& 20 	& 200 	& 603 	& 20 	& 200 	& 600 	\\ 
     $|R|$	& 4 	& 10 	& 43 	& 4 	& 10 	& 43	\\ 
     $|S|$	& 50 	& 300 	& 21945 & 50 	& 300 	& 21945	\\ 
     $|L_s|$	& 1-5 	& 1-6 	& 1-5 	& 2-6 	& 2-6 	& 2-6 	\\ 
     \begin{tiny}$\#$Maj\end{tiny}	& N/A 	& N/A 	& N/A 	& 4 	& 10 	& 20 	\\ 
     \textsc{SSS}	& $10^{182}$ & $10^{1840}$ & $10^{6695}$ & $10^{182}$ & $10^{1840}$ & $10^{6689}$ \\
    \hline  
    \end{tabular} 
	\end{small}
	\end{center}

	The large data set was created with parameters corresponding to the fall 2009 
	University of Toronto exam schedule, which is the size we expect for a large 
	real-world problem instance.

@@

  \subsection{Comparison Between the Different Searches}
    We tested the 4 search methods on all 6 of our problem instances. The results 
    for two of the problems, the small and large problems, are shown on figure 2 
    and 3. The medium problem exhibited characteristics very similar to the large problem.

@@

    \begin{figure}[h]
  	  \centering
  	  \includegraphics[width=8cm]{SMALL-TYPE2_FullAverageStats}
      \caption{Comparison of the different searches on a small data set}
	\end{figure}

	\begin{figure}[h]
  	  \centering
  	  \includegraphics[width=8cm]{HUGE-TYPE2_FullAverageStats}
      \caption{Comparison of the different searches on a large data set}
	\end{figure}

>>>>

  A schedule, then, is a mapping from courses to rooms at times, 
  which we can express like this:
  \[ C \rightarrow \left\{(r,\ a,\ b)\ |\ r\in R,\ a\in \{1..d\},\ b\in \{1..t\}\right\} \]
  Typically, with relatively loose constraints on the number of rooms and 
  particular schedules of students, it is not difficult to find some consistent
  schedule; that is, one where no student is asked to write 2 exams
  simultaneously and no room has 2 exams occur in it simultaneously.

@@

  The timetable that a student $s \in S$ has under some particular schedule 
  $K$ may be represented in this way:
  \[ TT_s(K) = \left\{(a, b)\ |\ \exists r \in R, \exists c \in L_s, K(c)=(r, a, b) \right\} \]
  which may, in general, be a multiset.

  Similarly, the timetable for a room $r \in R$ is
  \[ TT_r(K)=\left\{(a, b)\ |\ \exists c\in C,\ K(c)=(r, a, b)\right\} \]

  We define two quality functions, mapping schedules to real numbers in [0,1],
  one for students and one for rooms. These are meant to capture how much they
  ``like" the schedule under consideration. (For instance, a student will rate
  poorly any schedule where she has 2 consecutive exams, and very poorly any 
  schedule which expects her to take two exams at the same time. A room will 
  rate poorly a schedule that assigns two different exams to be held in it at 
  the same moment).
  \[ Q_s(K)=q_{student}\left(TT_s(K)\right):\textit{ Schedules}\rightarrow [0,\ 1] \]
  \[ Q_r(K)=q_{room}\left(TT_r(K)\right):\textit{ Schedules}\rightarrow [0,\ 1] \]
  We will consider later how to define these functions in a reasonable way.

  The quality of a schedule, then, is given by
  \[ Q(K)= \frac{\left(w_s\left(\sum_{s \in S}{\frac{Q_s(K)}{|S|}}\right) +
     w_r\left(\sum_{r\in R}{\frac{Q_r(K)}{|R|}} \right)\right)}{(w_s+w_r)} \]
  where $w_s$ and $w_r$ are the weightings we can use to indicate that we 
  care somewhat more about students preferences than rooms, giving a quality
  (or fitness, in the terminology of genetic algorithms) in the range [0,1] for
  any schedule.

@@

\section{Approach and Implementation}
  \subsection{Differences From Natural Exam Scheduling}
    The problem we have constructed here is a compromise between more mathematically
    pure problems with the same sort of properties (which might ignore the rooms' 
    constraints completely) and the problem that must be solved in the real world 
    when scheduling thousands of students, which has some important differences from 
    this problem. For example, we don't deal with constraints like room capacity or 
    exams with varying length.

@@

  \subsection{Fitness Function}
    In order to assign each schedule a fitness, we need to define the functions
    $q_{student}$ and $q_{room}$. They need to capture the idea of determining 
    how much each student or room likes their given timetable. We do this by giving 
    penalties to schedules when they have properties that are disliked by the 
    student or room, and then using the formula
    \[ q(TT)=\frac{1}{1+penalty_{TT}} \]
    This function will be in the desired range of [0,1].

@@

    \subsubsection{Representation}
      We use the naive representation of schedules. A schedule contains lists of size $|C|$
      containing the room in which each course's exam is to occur and the timing
      (corresponding to a pair or integers representing the day and time) of each course's
      exam. We will see that this representation is far from ideal, but there is no clear 
      alternative.

@@

      It can be seen that this method is completely deterministic given a starting schedule,
      and will always return a local maximum.

    \subsubsection{Mutate Search}
      Mutate Search starts by generating a random schedule. It makes an alternative schedule
      (by using the same mutate operator made for the genetic algorithm) with 1\% of the
      data randomized and keeps the better schedule. This operation is
      repeated until it has performed as many evaluations as it is allowed to.

      This method can be changed to true simulated annealing by varying the degree of
      modification made to the current best schedule with the number of evaluations
      remaining. Not being the focus of our investigation, little effort was made to 
      optimize the parameters. 
      %1\% does seem to give better performance than other nearby 
%      values, though we did not collect data to substantiate this. Finding ideal parameters 
 %     for simulated annealing on this problem would make for a project-sized inquiry in itself.

@@

    The large problem instance was created with parameters corresponding to the fall 2009 
	University of Toronto exam schedule, which is the size we expect for a large 
	real-world problem instance.

    \definecolor{light-gray}{gray}{0.87}
    \rowcolors{1}{light-gray}{white}
    \begin{table}[h]
    \begin{small}
    \begin{tabular}{*{7}{c}}
    \hline 
 	  	& S 	& M 	& L 	& $S_{Cor}$ & $M_{Cor}$ & $L_{Cor}$ \\ 
    \hline \hline
     d 		& 7 	& 10	& 7 	& 7 	& 10 	& 7 	\\ 
     t 		& 5		& 8 	& 8 	& 5 	& 8 	& 8 	\\ 
     $|C|$	& 20 	& 200 	& 603 	& 20 	& 200 	& 600 	\\ 
     $|R|$	& 4 	& 10 	& 43 	& 4 	& 10 	& 43	\\ 
     $|S|$	& 50 	& 300 	& 21945 & 50 	& 300 	& 21945	\\ 
     $|L_s|$	& 1-5 	& 1-6 	& 1-5 	& 2-6 	& 2-6 	& 2-6 	\\ 
     \begin{tiny}$\#$Maj\end{tiny}	& N/A 	& N/A 	& N/A 	& 4 	& 10 	& 20 	\\ 
     \textsc{SSS}	& $10^{182}$ & $10^{1840}$ & $10^{6695}$ & $10^{182}$ & $10^{1840}$ & $10^{6689}$ \\
    \hline  
    \end{tabular} 
	\end{small}
	\caption{The six problem instances used in our project. $\#$Maj is the number of different 
	         majors used and \textsc{SSS} is an approximation of the search space size.}
	\end{table}

@@

  \subsection{Comparison Between the Different Searches}
    We tested the 4 search methods on all 6 of our problem instances. The results 
    for two of the problems, the small and large problems, are shown in figure 2 
    and 3. The medium problem exhibited characteristics very similar to the large problem.

@@

    \begin{figure}[h]
  	  \centering
  	  \includegraphics[width=8cm]{SMALL-TYPE2_FullAverageStats}
      \caption{Search methods on small problem instance}
	\end{figure}

	\begin{figure}[h]
  	  \centering
  	  \includegraphics[width=8cm]{HUGE-TYPE2_FullAverageStats}
      \caption{Search methods on large problem instance}
	\end{figure}
